Normalization by Evaluation for Call-by-Push-Value and Polarized Lambda-Calculus

### 2019-02-14 fscd19/fscd19.tex
[ fscd19 ] start on introduction

The idea behind \emph{normalization by evaluation} (NbE) is to utilize a
standard interpreter, usually evaluating closed terms, to compute the
normal form of an open term.

Weak coproducts.
Weak sheaf (non-unique pasting).

Contributions:
\begin{enumerate}
\item Abstract cover monad.
\item NbE for CBPV and polarized lambda-calculus.
\end{enumerate}

CBPV differs from PolLambda in that hypotheses can be arbitrary
positive formulas in CBPV, but only positive atoms or negative
formulas in PolLambda.  This means that in CBPV, there is more choice
on when and how to decompose a hypothesis into ones of smaller
positive type, whereas in PolLambda there is no freedom: mandatory
maximal splitting is dictated by the left inversion phase.

Normal forms of CBPV thus retain individual shallow pattern matching
constructs, while PolLambda has only deep, maximal pattern matching.

Agda formalization:
Sized types, nested data types.

Structural focalization for sequent calculus \cite{simmons:tocl14}.

\nocite{bergerSchwichtenberg:lics91}

\nocite{jungTiuryn:tlca93}

\nocite{altenkirchHofmannStreicher:ctcs95}

\nocite{abel:habil}

\nocite{danvy:popl96}

\nocite{fioreSimpson:tlca99}

\nocite{altenkirchDybjerHofmannScott:lics01}

\nocite{balatDiCosmoFiore:popl04}

\nocite{altenkirchUustalu:flops04}

\nocite{barral:PhD}

\nocite{scherer:popl17}

@@

\section{Focused Intuitionistic Propositional Logic}
\label{sec:fipl}

>>>>

The idea behind \emph{normalization by evaluation} (NbE) is to utilize a
standard interpreter, usually evaluating closed terms, to compute the
normal form of an open term.
The normal form is achieved by a type-directed \emph{reification}
procedure after evaluating the open term to a semantic value, mapping
the free variables to corresponding \emph{unknowns} in the semantics.
The literal use of a standard interpreter
can be achieved for the pure simply-typed lambda-calculus
\cite{bergerSchwichtenberg:lics91,filinski:semaccounttdpe}
by modelling uninterpreted base types as sets of neutral (aka atomic)
terms, or more precisely, as presheaves or sets of neutral term
families, in order to facilitate fresh bound variable generation during
reification of functions to lambdas.
Thanks to $\eta$-equality at function types, free variables of
function type can be reflected into the semantics as functions
applying the variable to their reified argument,
becoming a neutral.  This mechanism provides us with unknowns of
function type which can be faithfully reified to normal forms.
% Since neutrals are closed under application to normal arguments, they
% can act as unknown functions, storing the reified semantic argument
% they are invoked with (thus, these unknowns are aka accumulators
% \cite{gregoireleroy:icfp02}).

For the extension to sum types (logically disjunctions and categorically
weak coproducts), this reflection trick does not work anymore.  A
semantic value of binary sum type is either a left or a right
injection, but the decision between left or right cannot be taken at
reflection time, since a variable of sum type does not provide us with
the information.  A literal standard interpreter for closed terms can
thus no longer be used for NbE.  Instead, we can utilize a monadic
interpreter.  When the interpreter attempts a case distinction on an
unknown of sum type, it asks an oracle whether the unknown is a left
or a right injection.  The oracle returns one of these alternatives,
wrapping a new unknown in the injection.  The communication with the
oracle can be modeled in a \emph{monad} $\C$,
% with the service $\tNe\,(A + B) \to \C\,(\den A + \den B)$
which records the questions asked and the continuation of the
interpreter for each of the possible answers.
A monadic semantic value is thus a \emph{case tree} where the leaves
are occupied by non-monadic values.
In this article we only consider weak sum types, producing non-unique
normal forms, where it does not
matter in which order the questions are asked (commuting case splits),
and whether the same
question is asked several times (redundant case splits).
The model would need refinement for strong, extensional sums
\cite{altenkirchDybjerHofmannScott:lics01,altenkirchUustalu:flops04,balatDiCosmoFiore:popl04,barral:PhD,scherer:popl17}.

Filinski \cite{filinski:tlca01} studied NbE for Moggi's computional lambda
calculus \cite{moggi:infcomp91} shedding light on the difference
between call-by-name (CBN) and call-by-value (CBV) NbE, where Danvy's
type-directed partial evaluation \cite{danvy:popl96} falls into the
latter class.  The contribution of the computational lambda calculus
is to make explicit where the monad is envoked during monadic
evaluation, and this placement of the monad carries over to the NbE
setting.  Moggi's studies were continued by Levy \cite{levy:hosc06}
who designed the call-by-push-value (CBPV) lambda-calculus to embed
both the CBN and CBV lambda calculus.

In this work, we formulate NbE for CBPV
(Section~\ref{sec:cbpv}), with the aim to investigate
later whether CBN and CBV NbE can be recovered from CBPV NbE via the
standard translations of the CBN and CBV calculi into CBPV.

In contrast to the normal forms of CBN NbE, which is the algorithmic
counterpart of the completeness proof for intuitionistic propositional
logic (IPL) using Beth models, CBPV NbE gives us more restrained
normal forms, where the production of a value via injections cannot be
interrupted by more questions to the oracle.  In the research field of
focalization \cite{andreoli:focusing,liangMiller:csl07} we speak of
\emph{chaining non-invertible introductions}.
Invertible introductions are already chained in NbE thanks to
extensionality ($\eta$) for function (an more generally, negative)
types.  Non-invertible eliminations are also happing in a chain,
building neutrals.  What is missing from the picture is the chaining
of invertible eliminations, i.e., case distinctions and, more
generally, pattern matching.  The picture is completed by extending
NbE to \emph{polarized lambda calculus}
\cite{zeilberger:PhD,brockNannestadSchuermann:lpar10,espiritoSanto:entcs17}
in Section~\ref{sec:fipl}.

In our presentation of the various lambda calculi and NbE, we use
presheaves for the abstract syntax and the semantics
\cite{altenkirchHofmannStreicher:ctcs95}.

In contrast to Zeilberger \cite{zeilberger:PhD}.

% Weak coproducts.
% Weak sheaf (non-unique pasting).

% Contributions:
% \begin{enumerate}
% \item Abstract cover monad.
% \item NbE for CBPV and polarized lambda-calculus.
% \end{enumerate}

% CBPV differs from PolLambda in that hypotheses can be arbitrary
% positive formulas in CBPV, but only positive atoms or negative
% formulas in PolLambda.  This means that in CBPV, there is more choice
% on when and how to decompose a hypothesis into ones of smaller
% positive type, whereas in PolLambda there is no freedom: mandatory
% maximal splitting is dictated by the left inversion phase.

% Normal forms of CBPV thus retain individual shallow pattern matching
% constructs, while PolLambda has only deep, maximal pattern matching.

% Agda formalization:
% Sized types, nested data types.

% Structural focalization for sequent calculus \cite{simmons:tocl14}.

% \nocite{bergerSchwichtenberg:lics91}

% \nocite{jungTiuryn:tlca93}

% \nocite{altenkirchHofmannStreicher:ctcs95}

% \nocite{abel:habil}

% \nocite{danvy:popl96}

% \nocite{fioreSimpson:tlca99}

% \nocite{altenkirchDybjerHofmannScott:lics01}

% \nocite{balatDiCosmoFiore:popl04}

% \nocite{altenkirchUustalu:flops04}

% \nocite{barral:PhD}

% \nocite{scherer:popl17}

@@

\section{Focused Intuitionistic Propositional Logic}
\label{sec:fipl}
\label{sec:pol}
